### 2020-08-05 paper/sex_selection_spacing-ver4.tex
Edit birth interval section

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant 
changes in India.
Figures \ref{fig:spacing_low_rural} through \ref{fig:spacing_highest_urban}
show the 25th, 50th, and 75th percentiles birth intervals in 
months, the sex ratio, and the probability of having a birth for each spell
for the four education levels by area of residence.%
\footnote{
Results for urban women without education, rural women with twelve or more years of 
education, and the fourth spell for women with twelve or more years of education are
presented in the online appendix because they are based on relatively few observations.
}
The underlying values with bootstrapped standard errors are available in the online 
appendix.
The sex ratio graphs also show the natural sex ratio, approximately 51.2 percent 
\citep{ben-porath76b,jacobsen99,Portner2015b}.

% [not sure where this goes or simply delete]
%Sex selection increases all three percentile birth intervals, but the effects of abortions 
%are easiest to observe in the 75th percentile birth interval, 
%and it is also the changes in the longest birth intervals that will likely the largest 
%impact on fertility measures.
%The 25th percentile captures changes in the short spacing, which is the most likely to 
%have an adverse impact on health.

@@

% [Parity progression probability]
The parity progression and the sex ratio show two broad trends.
First, 
in line with the observed fall in the total fertility rate, the likelihood of a next birth 
has fallen over time.
The declines in the likelihood of a next birth are larger the higher the education, 
the higher the parity, and with the presence of at least one son. 
Within a given spell and period, parity progressions are lower in urban than in rural areas, 
if at least one son is present, and the more educated the mother.

% [Sex ratios]
Second, 
the spread of sex selection shows clearly in the sex ratios of next births, which has 
become more male dominated for women with no sons.
The increases in the predicted percent boys are larger the higher the education and the 
higher the parity.
There are no clear trends for the other sex compositions.
Furthermore, for a given spell and period, sex ratios in the absence of a son are higher 
in urban than in rural areas, the higher the education of the mother, and the higher
the parity.
There is some evidence of unnaturally high percentage boys for some women with one 
son, although most are clearly lower than in the absence of a son and the failing fertility
make these estimate noisy.

% [Birth intervals w/o sex selecition]

To separate out the effects of the introduction of sex selection and the other changes in
India, I first discuss how birth intervals have changed in situations where sex selection are 
less likely to be used.
This groups covers women with no education, no matter the sex composition of their children, 
and women with any education who has one or two sons already. 
[This is not a hard and fast separation ]

% [median birth intervals]
Most median birth intervals have increased by three to six months over the four 
decades, but the majority of median birth intervals are still at 36 months or below, 
with the shortest only 29 months.%
\footnote{
The NFHS reports show median closed birth intervals of approximately 31 months, which have 
barely moved over time, underscoring the importance of accounting for censoring when examining 
birth spacing.
}
These changes are substantially smaller than the 3.5 months increase per decade found
in other countries with declining fertility \citep{Rutstein2011,Casterline2016}.

% [25th percentile birth intervals]
Despite the increases in median birth intervals and the decline in fertility, a remarkably 
high proportion of birth intervals are still very short.
For all but the most educated, more than 25 percent of women have their second and their
third child within 24 months of their previous child, which is substantially below
the 24 between \emph{pregnancies} recommended by the WHO.
Furthermore, even with the larger increases in birth intervals with higher parity, the 
25th percentile birth intervals for the fourth spell is around 24 months for women with 
less than eight years of education.

% [75th percentile birth intervals]
The most substantial differences across women come from the longer birth intervals, 
captured by the 75th percentile birth intervals.
For the two lowest education groups the increases in the 75th percentile birth 
intervals are relatively small for the second spell and slightly more substantial for the 
third spell at around five months.
However, the fourth spell show significant increases, especially for women with two
or three sons, with increases of fifteen to twenty months.

These results are in line with prior research showing that falling fertility is 
associated with increases in longer spacing, although it is still an 
unresolved question why \citep{Casterline2016}.
The parity progression probabilities have been falling over time across all 
education levels, especially for women with at least one son, and have been
failing less for the less educated women.

However, despite the rapidly declining probability of a third birth for the best-educated
women when they already have a son, the birth intervals changed little for any of the percentiles,
and all are lower for the third spell than for the second.
One possibility is that the most educated women have both better access to modern
contraceptive and are better at using traditional contraceptive methods 
\citep{Rosenzweig1989}.

@@

Women with the largest increases in birth intervals are also those least likely to work.
For example, the increases in average third spell birth intervals for urban women with
one boy and one girl over the four decades were 2.7, 3.4, 5.8, and 1.8 months from lowest 
to highest education groups.%
\footnote{
For women with two sons, the numbers were 4.3, 6.3, 7.0, and 3.0, respectively.
See the online appendix tables for the average birth intervals.
}
This means that the group with the lowest labor force participation, women with eight to 
eleven years of education, was also the group with the largest increases in average 
spacing, likely driven by the substantial improvement in household income that came
from economic growth.

Son preference still show up clearly, even for the women the least likely to use sex 
selection, such as rural women with no education. 
For this group, the shortest spacing---across all spells---is when they have only girls. 
Notably, the difference in birth intervals across sex compositions has grown over 
time as spacing when sons are present increased.

% [Birth intervals with sex selection]

The clearest illustration of how the combination of son preference and the introduction of 
sex selection affected birth spacing comes from the third spell of the best-educated urban 
women.
Women with two girls have a predicted sex ratio for their third child approaching 80 
percent boys and a 75th percentile birth interval of almost 70 months, about 13 months
longer than if they had at least one son, and an increase of more than 20 months over
the four decades.

Even more striking is the change for this group when sex selection was first introduced.
The 75th percentile birth interval with two girls increased from 48 months to 
64 months with two girls in a decade, while the other sex compositions showed a 
slightly decrease from around 55 months to 54 months.
Theses changes may even be an underestimate of the changes in birth spacing since
this particular group appear to have had access to sex selection even before it became
widespread as shown by the unequal sex ratio for the 1972--1984 period for women with
two girls.

Not surprisingly given these effects of sex selection, the third spell for the 
best-educated women is also the case with the clearest reversal in the spacing pattern by 
sex composition; the birth intervals with two girls are consistently longer than with one
or two sons no matter the percentile used.
[highlight others]

@@

Did the predictions of declining use of sex selection come true?
There is no clear evidence for or against a reversal in the use of sex selection, with 
some cases showing increases in sex ratios between the last two periods, others little change, 
and some a decline.
The best-educated women is again a good illustration.
The percent boys born to women with two girls continued to increase over the last two
periods, but the likelihood of a third birth declined and when the first child was a girl, 
the sex ratio for the second birth dropped slightly as did the probability of a second birth. 
There is, however, also cases where there is clearly no abatement in the increasing use of 
sex selection. 
The sex ratios in the absence of girls continued to increase for rural women with
one to seven years of education and the likelihood of an additional birth remained high.

Finally, although birth spacing has increased over time with improving economic 
conditions and falling fertility, most of the increases in birth spacing come from the
increased use of sex selection. 
The most substantial increase over the four decades is a full year for the best-educated 
women with two girls, which is the group with the most biased sex ratio.

First, 
over the four decades, birth intervals have lengthen, whether measured by the
25th, 50th, or 75th percentile.
These increases are larger the higher the education of the mother, the higher the parity, 
and the higher the percentile measure.

>>>>

% Explain question and what I am doing
The first question I address is how birth spacing responded to the significant 
changes in India.
Figures \ref{fig:spacing_low_rural} through \ref{fig:spacing_highest_urban}
show the 25th, 50th, and 75th percentiles birth intervals in 
months, the sex ratio, and the probability of having a birth for each spell
by education levels and area of residence.%
\footnote{
Results for urban women without education, rural women with twelve or more years of 
education, and the fourth spell for women with twelve or more years of education are
in the online appendix because of relatively small samples.
}
The sex ratio graphs also show the natural sex ratio, approximately 51.2 percent 
\citep{ben-porath76b,jacobsen99,Portner2015b}.
The underlying values with bootstrapped standard errors are available in the online 
appendix.

@@

% [Parity progression probability]
The parity progression and the sex ratio show two broad trends.
First, 
in line with the falling total fertility rate, the likelihood of a next birth 
has decreased over time.
The likelihood of a next birth fell faster, the higher the education, the higher the 
parity, and with at least one son.
Within a given spell and period, parity progressions are lower in urban than in rural areas, 
if at least one son is present, and the more educated the mother.

% [Sex ratios]
Second, 
the spread of sex selection shows clearly in the sex ratios of next births, which has 
become more male-dominated for women with no sons.
The percent of births that were boys increased faster, the higher the education and the 
higher the parity.
There are no clear trends for the other sex compositions.
Within a given spell and period combination and in the absence of a son, sex ratios are 
higher, the more educated the mother, and the higher the parity. 
Sex ratios are also higher in urban than in rural areas.
Some women with one son also show an unnaturally high percentage of boys, although the 
failing fertility makes these estimates noisy.

% [Birth intervals w/o sex selection]
\subsection{When Sex Selection is Less Used}

To separate the effects of the introduction of sex selection and the other changes in
India, I first discuss how birth intervals have changed in situations where sex selection 
is less used.
The group broadly covers women with no education, no matter the sex composition of their 
children, and women with any education who has one or two sons already. 
Despite the lower level of sex selection, son preference is still evident with the 
shortest spacing when they have only girls. 
Notably, for those least likely to use sex selection, rural women with no education, the 
difference in birth intervals across sex compositions has grown over time as spacing when 
sons are present increased.

% [25th percentile birth intervals]
A remarkably high proportion of birth intervals are still very short.
For all but the most educated, more than 25 percent have their second and third child 
within 24 months of the previous birth, substantially below the 24 months between 
\emph{pregnancies} the WHO recommends.
Furthermore, despite higher parities' more substantial increases in birth intervals, even 
the 25th percentile birth intervals for the fourth spell are around 24 months for women 
with less than eight years of education.

% [median birth intervals]
Median birth intervals have also increased relatively little---only three to six months 
over the four decades---compared to around 3.5 months \emph{per decade} in other countries with 
declining fertility \citep{Rutstein2011,Casterline2016}.%
\footnote{
The NFHS reports show median closed birth intervals of approximately 31 months, which have 
barely moved over time, underscoring the importance of accounting for censoring when examining 
birth spacing.
}
The result is that most of the median birth intervals are still at 36 months or below, 
with the shortest only 29 months.

Birth intervals appeared to lengthen the most for women the least likely to work.
For example, from lowest to higher education, the average third spell birth intervals for 
urban women with one boy and one girl increased by 2.7, 3.4, 5.8, and 1.8 months over the 
four decades.%
\footnote{
For women with two sons, the numbers were 4.3, 6.3, 7.0, and 3.0.
See the online appendix tables for the average birth intervals.
}
Hence, women with the lowest labor force participation, those with eight to eleven years 
of education, also saw the largest increases in average spacing, possibly driven by the 
substantial improvement in household income for this group from economic growth.

% [75th percentile birth intervals]
The most substantial changes occurred in the 75th percentile birth intervals, where the 
more the parity progression probabilities declined, the more the birth interval lengthen. 
For example, the probability of a fourth birth for urban women with eight to eleven years 
of education and two sons and a girl has declined by almost forty percentage points as the 
75th percentile birth interval increased by 22 months.
Compare this with no education, rural women with a boy as their first child, where the 
probability of a third birth declined by less than six percentage points while the birth 
interval increased by only slightly over two months.

%0.637 0.241

These results are in line with prior research showing that falling fertility is associated 
with increases in longer spacing, although why is still an unresolved question 
\citep{Casterline2016}.
The exception to this trend suggests one possible answer.
For the most educated women who already have a son, the probability of a third birth 
declined rapidly, but the birth intervals changed little.
These women have both better access to modern contraceptives and are better at using 
traditional contraceptive methods \citep{Rosenzweig1989}.

@@

% [Birth intervals with sex selection]
\subsection{Sex Selection and Birth Spacing}

A clear illustration of how the combination of son preference and the introduction of 
sex selection affected birth spacing comes from the third spell of the best-educated urban 
women.
With two girls, almost 80 percent of the third births are boys, and the 75th percentile 
birth interval is close to 70 months. 
This interval is about 13 months longer than if they had at least one son and represents 
an increase of almost 21 months over the four decades.

Even more striking is that most of the change took place right at the introduction of sex 
selection.
The 75th percentile birth interval with two girls increased from 48 months to 
64 months in a decade, while the other sex compositions showed a slight decrease from 
around 55 months to 54 months.
These changes in birth spacing may even be an underestimate since this particular group 
appears to have had access to sex selection even before it became widespread, as shown by 
the unequal sex ratio for the 1972--1984 period for women with two girls.

The 75th percentile changes are the most dramatic, but sex selection also impacts the 
25th and median birth intervals. 
For the best-educated urban women with two girls, the 25th percentile birth interval 
increased by six months, or 23 percent, while the median percentile birth interval 
increased by 15 months (43 percent).

Not surprisingly, given these effects of sex selection, the third spell for the 
best-educated women shows the clearest reversal in the spacing pattern; 
the birth intervals with two girls are consistently longer than the intervals with one or 
two boys, no matter the percentile used.
A similar reversal, although more muted, occurred for the third spell for urban women 
with one to seven years of education and both urban and rural women with eight to
eleven years of education.

@@

Did the predictions of declining use of sex selection come true? 
There is no clear evidence for or against a reversal in the use of sex selection, with 
some cases showing increases in sex ratios between the last two periods, others little 
change, and some a decline. 
The best-educated women are again a good illustration. 
The sex ratio for women with two girls continued to increase over the last two periods, 
but the likelihood of a third birth declined. 
Furthermore, if the first child was a girl, the sex ratio for the second birth dropped 
slightly, as did the probability of having a second birth. 
However, there are also cases where there is no abatement in the increasing use of sex 
selection. 
For example, for rural women with one to seven years of education, the sex ratios in the 
absence of girls continued to increase while the likelihood of an additional birth remained 
high.

In summary, over the four decades, birth intervals have lengthened with improving economic 
conditions and falling fertility. 
These increases are larger with higher parity and higher percentile measure. 
Furthermore, when sex selection is less used, it appears that the women least likely to 
work are also those with the most substantial increases.

Sex selection, however, is behind the most substantial increases in birth spacing. 
The best-educated women with two girls had the most biased sex ratio and the most 
significant increase in birth intervals. 
Over the four decades, the median birth interval for this group increased by almost 15, 
and the 75th percentile birth interval increased by a staggering 21 months, most of that 
within a decade.
